---
title: "Technical comment on Evidence for a limit to human lifespan"
author: "Philipp Berens"
date: "October 9, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(plyr)
library(rstanarm)

```

Dong et al. claim to present statistical evidence in favor of an absolute limit to the human lifespan. Here I present a reanalysis of a central figure in their paper showing that in fact the data is uninformative with regards to the question whether there is a limit to human lifespan or not. 

## Figure 2a

The authors graph the maximum age reported at death for each year between 1968 and 2006. I acquired the data using WebPlotDigitizer and rounded the numbers to full years (which is what likely was the case for the original data). 

Here is the raw data, as presented by the authors, fitting separate regression for years up to 1994 and after 1995.  


```{r, echo=FALSE}
tbl <- read.csv('lifeexpectancy.csv')
tbl <- round(tbl)
tbl$Group <- factor(tbl$Year>=1995, levels = c("FALSE", "TRUE"), labels = c("<1995", ">=1995"))


ggplot(tbl,aes(x=Year,y=Age,Color=Group)) + geom_point() + 
  geom_smooth(method="lm") + 
  ylab('Yearly maximum reported age at death (years)')  


```

## A simple, linear model

We first fit a simple linear model to the data:

```{r, echo=FALSE}

ggplot(tbl,aes(x=Year,y=Age)) + geom_point() + 
  geom_smooth(method=lm) + 
  ylab('Yearly maximum reported age at death (years)')  

```

The plots shows the raw data points again, with a linear regression fitted to all the data as well as 95% confidence intervalls. 

```{r}

mdl1 <- lm(Age~Year,tbl)
summary.lm(mdl)

```

The classical frequentist model fit shows a significant overall p-value, and a significant influence of the factor Year. The regression reveals that the maximum reported age at death (MRAD) increases slightly by `r signif(mdl$coefficients['Year'],digits = 2)` years or  `r signif(mdl$coefficients['Year'],digits = 1)*12` months. 

When inspecting the model fit, we find that a couple of points appear to be systematically missed by the linear model: 


## Two group model fitted by the authors

```{r, echo=FALSE}

ggplot(tbl,aes(x=Year,y=Age)) + geom_point() + 
  geom_smooth(data=tbl,method=lm,formula = Age~Year*Group) + 
  ylab('Yearly maximum reported age at death (years)')  

```




```{r}

mdl2 <- lm(Age~Year*Group,tbl)
summary.lm(mdl2)

tbl$AgePred2 <- predict.lm(mdl2, tbl$Year)

ggplot(tbl,aes(x=Year,y=Age)) + geom_point() + 
  geom_line(data=tbl,aes(x=Year,y=AgePred2)) + 
  geom_line(data=tbl,aes(x=Year,y=AgePred1)) + 
  
  ylab('Yearly maximum reported age at death (years)')  


BIC(mdl1)
BIC(mdl2)

```

difference not worth mentioning


##Robust Regression
```{r}

rmdl1 <- rlm(Age~Year,tbl)
summary(rmdl1)

rmdl2 <- rlm(Age~Year*Group,tbl)
summary.lm(rmdl2)

tbl$AgePred2 <- predict.lm(mdl2, tbl$Year)

ggplot(tbl,aes(x=Year,y=Age)) + geom_point() + 
  geom_line(data=tbl,aes(x=Year,y=AgePred2)) + 
  geom_line(data=tbl,aes(x=Year,y=AgePred1)) + 
  
  ylab('Yearly maximum reported age at death (years)')  

```



##Bayesian modeling

```{r, cache=TRUE}

bmdl1 <- stan_lm(Age~Year,tbl,prior = R2(location = .5, what="mean"))
summary(bmdl1)


bmdl2 <- stan_lm(Age~Year*Group,tbl,prior = R2(location = .5, what="mean"))
summary(bmdl2)

loo_bmdl1 <- loo(bmdl1, k_threshold = 0.7)

loo_bmdl2 <- loo(bmdl2, k_threshold = 0.7)

compare(loo_bmdl1,loo_bmdl2)


```

Reference https://arxiv.org/abs/1507.04544







